---
title: "Economic models and calibration"
output: rmarkdown::html_vignette
bibliography: antitrustbib.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Economic models and calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

To predict the effect of a merger, we need to know where sales go when a
price changes and how firms take those sales into account. Prices and shares
alone rarely answer either question. `antitrust` combines observed market
information with a demand system and an assumption about firm behavior to
recover the parameters needed for a simulation.

The answer is conditional on those inputs and assumptions. A different
margin, diversion estimate, or conduct model can change recovered demand and
costs, and therefore the predicted effect. Calibration does not provide
sampling standard errors; checking plausible inputs and model choices is
part of an economic analysis. The [workflow vignette](Workflow.html) shows
how to use a fitted market. Here we explain what its fit means.

## From observed market to structural state

Most calibrations start with product prices $p$, quantities or shares $q$,
pre-change ownership $\Omega$, and at least some gross margins
$m_i=(p_i-c_i)/p_i$. A model may also require diversion ratios,
elasticities, capacities, bargaining powers, nest assignments, or auction
moments. Check what a "share" means before supplying one: Logit uses
*quantity* shares, while CES and AIDS use *revenue* shares. Utilities and
valuations may also need an observed outside share or a normalization to fix
their common level.

Consider Bertrand pricing. Firm $f$ chooses the prices of its products to
maximize $\sum_{j\in J_f}(p_j-c_j)q_j(p)$. For one of those products, $i$,
the first-order condition is

$$
q_i(p)+\sum_{j\in J_f}(p_j-c_j)\frac{\partial q_j(p)}{\partial p_i}=0.
$$

The first term is the gain from charging a slightly higher price on units of
$i$ still sold. The sum captures lost sales of $i$ and any change in profits
on the firm's other products. A merger adds terms to that sum: sales
diverted from one merging product to another now matter to the same owner.
Partial ownership is handled with ownership weights.

Observed shares and margins restrict the demand derivatives in this
condition. The chosen demand system fills in the remaining structure and
allows marginal costs to be recovered. A counterfactual then changes
ownership or another market condition while holding the calibrated demand
primitives fixed. Two demand systems can agree at the observed prices but
predict different outcomes if their derivatives change differently away from
that point.

One useful observation about substitution is quantity diversion from $i$ to
$j$:
$D^q_{ij}=-(\partial q_j/\partial p_i)/(\partial q_i/\partial p_i)$.
It can help identify derivatives that shares and margins leave
undetermined. Under the package's diversion-matrix convention, the diagonal
is $-1$. With an outside option, the other modeled products need not absorb
all sales lost by $i$. See `?diversion` for each model's quantity or revenue
convention.

## Demand systems

Demand assumptions determine both **who substitutes for whom** and **how
those substitution patterns change** as prices move. Linear and Log-linear
models put restrictions directly on slopes or elasticities. Logit and CES
give substitution a more structured form, while nests and random
coefficients allow more variation. The table gives the main calibration
tradeoffs; the model help pages give exact arguments and solver controls.

| Family | Substitution and recovered parameters | What the calibration needs or restricts |
|:--|:--|:--|
| Linear | $q=a+Bp$ has constant slopes in quantity levels. Calibration recovers intercepts and a slope matrix. | Prices, quantities, margins, and diversion or symmetry restrictions are needed to pin down slopes. There is no curvature in quantity levels. |
| Log-linear (`LogLin`) | $\log q_i=a_i+\sum_j B_{ij}\log p_j$ has constant elasticities, collected in $B$. | Prices, quantities, margins, and diversion or other restrictions identify the elasticities. |
| LA-AIDS (`AIDS`) | Budget shares respond to log prices and an expenditure/price index. Adding up, homogeneity, and symmetry restrict cross-price effects. | Revenue shares, margins, and diversion or curvature restrictions identify the local system. Prices can be optional at calibration but matter for level outputs. |
| PCAIDS and nested PCAIDS | Share proportionality restricts substitution; nests allow stronger within-group substitution. | `knownElast` and `mktElast` provide identifying information. Nested calibration also needs nests and enough margins. |
| Logit | $s_i=\exp(\delta_i+\alpha p_i)/(1+\sum_j\exp(\delta_j+\alpha p_j))$. A common $\alpha<0$ governs price response; substitution is proportional to shares. | Prices, quantity shares, ownership, and enough margins identify $\alpha$ and mean values. Without an outside option, one mean value is normalized. |
| Nested Logit | Correlation within nests directs more diversion toward products in the same nest. | Nest assignments and additional margins identify nesting parameters. A singleton nest cannot identify its own parameter. |
| CES | Quality weights and price powers determine revenue shares. `gamma` controls substitution; `alpha` and the inside share govern expenditure allocation. | Prices, revenue shares, margins, and an outside share or normalization identify primitives for the chosen construction. Ordinary output-market translations require `gamma > 1`. |
| Nested CES | Substitution can differ within and across nests. | Nest assignments and extra restrictions or margins identify the curvatures. In ordinary output markets, within-nest `sigma` exceeds between-nest `gamma > 1`. |
| BLP/random coefficients | Integrating heterogeneous price tastes over consumer draws allows flexible diversion. | Prices, shares, known outside share `s0`, and conduct-specific margins identify price-taste parameters by contraction and margin matching. Calibration does not estimate `s0` or bargaining power. |
| Capacity Logit (`LogitCap`) | Logit demand is subject to product quantity ceilings. | Market size and capacities locate binding constraints. Margins at exactly binding products matter because unconstrained FOCs do not identify their costs. |

In Linear, Log-linear, and AIDS models, restrictions on derivatives and
diversion help fill gaps in observed market data [@Haefen2002;
@LaFrance2004; @Epstein2004]. Logit, CES, and capacity models instead put
more structure on substitution or supply constraints [@Werden1994;
@Sheu2011; @Froeb2003].

Some models have an `"alm"` variant for an unobserved outside share. It uses
additional margins to identify that share along with demand curvature. The
variant is a separate implemented model, even when its demand name is the
same, and it may lack a `specify()` route.

### Identification in observed synthetic markets

The observed synthetic experiment begins with shares, ownership, marginal
costs, and one proportional reference margin. One margin identifies a common
Logit price coefficient or CES curvature under the listed standard conduct
equations, but it does not identify every model in this table. Nested Logit
also requires the caller to supply `nests` and named `parameters$sigma`;
Linear Bertrand requires a relative `parameters$slopes` matrix. Bargaining
power defaults to 0.5 per product unless `bargpowerPre` is supplied.
PCAIDS Bertrand uses diversion proportional to share and aggregate elasticity
`-1`; the reference margin identifies its remaining demand scale. In ALM
Logit and CES models, the active-product shares sum to one conditional on
participation. A separate passive outside share is supplied or drawn
uniformly from 0.1 to 0.5. This supports Bertrand, Cournot, second-score
auction, and bargaining observed inversions for both demand families.
Nested Logit ALM Bertrand additionally uses supplied `nests` and named
`parameters$sigma` for the within-nest substitution pattern.

The registry's `observed_synthetic` column names implemented inversion
handlers; `"unsupported"` means the route requires more anchors or different
design state. The model-by-model reasons are recorded in
`artifacts/observed_synthetic_market_redesign.md`. Calibration from observed
prices remains available through `calibrate()` for other models, which is a
different experiment.

## Conduct and cost recovery

Demand tells us how buyers respond to prices; conduct tells us which choices
firms make and which profits they internalize. The distinction matters for
cost recovery as well as merger predictions. A Bertrand margin, for example,
cannot simply be fed into a Cournot first-order condition. Each conduct model
below uses its own equilibrium argument.

| Conduct | How firms compete and what a merger changes | What calibration requires |
|:--|:--|:--|
| Bertrand | Firms choose prices and internalize the margins on their own products. Common ownership adds recaptured sales to that calculation. | Margins help identify demand derivatives; fitted prices and margins yield costs. |
| Cournot | Firms or plants choose quantities, taking account of their effect on inverse demand. Common ownership changes those quantity incentives. | General Linear/LogLin implementations use plant quantities and production or cost information. Logit and CES Cournot have separate complete implementations. |
| Stackelberg | Quantity leaders move before followers, whose responses they anticipate. A merger or a capacity/leadership change may alter that sequence. | Implemented Linear/LogLin games require plant and leadership information. |
| Monopolistic competition (`moncom`) | Each product is priced against its *perceived own* demand. CES holds the aggregate index fixed for that derivative. | Prices, shares, and at least one margin identify supported Logit/CES/BLP variants. Ownership does not enter the cross-product FOC. |
| Differentiated second-score auction | A seller offers the portfolio product with the highest surplus and bids cost; payment reflects the next-best score. A merger changes portfolio alternatives. | Shares and margins recover offer-value and cost primitives. A valuation may need normalization; price data help separate value from cost under cost changes [@Brannman2000; @miller2014]. |
| Nash bargaining | Buyer and seller negotiate against disagreement payoffs. A merger changes recapture within the seller's portfolio and the bargaining alternatives. | Bargaining power is supplied. Supported Logit, CES, and BLP models use their own FOCs rather than Bertrand margins alone. |
| Second-score bargaining | Bargaining occurs with a second-score auction disagreement option. | The Logit and CES implementations each have their own demand, bargaining, and auction state. |
| Capacity-constrained second-price auction | A buyer sets a reserve, while seller costs are drawn from a distribution; capacity changes the minimum-cost distribution. | Capacities, prices/margins, and sometimes reserve or self-supply moments identify the cost distribution. More flexible distributions need more moments [@WaehrP2003]. |

## Implemented combinations

The demand and conduct descriptions above are a guide to choosing assumptions,
not a menu of components that can be paired at will. The table below is
generated from `supportedModels()` when this vignette is built. Each row is a
**complete implemented model**. `TRUE` means the listed lifecycle operation
is available; `simulate()` still checks whether a particular scenario field
is supported by that model.

```{r support-table, echo=FALSE}
models <- antitrust::supportedModels()
knitr::kable(
  models[c("demand", "conduct", "variant", "class",
           "calibrate", "specify", "simulate")],
  row.names = FALSE
)
```

The model registry identifies what can be fitted or simulated. A separate,
narrower transition registry determines which fitted models can be translated
with `respecify()`; an implemented row does not imply a transition from every
other row. The [workflow vignette](Workflow.html) shows such comparisons, and
the [architecture vignette](Architecture.html) explains the registry's role in
dispatch and extensions.
