---
title: "Economic models and calibration"
output: rmarkdown::html_vignette
bibliography: antitrustbib.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Economic models and calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`antitrust` fits **complete economic models**, each with a particular demand
system, strategic interaction, and identification argument. These models use
observed market moments and economic restrictions to recover structural
parameters; `calibrate()` is not a statistical estimator with standard errors.
Changing an observed margin or an assumed conduct model can change both
recovered demand and recovered costs. The [workflow vignette](Workflow.html)
shows how to use a fitted market; this vignette explains what is fitted.

## From observed market to structural state

The usual inputs are product prices $p$, quantities or shares $q$, pre-change
ownership $\Omega$, and some observed margins $m_i=(p_i-c_i)/p_i$.
Diversion ratios, elasticities, capacities, bargaining powers, nests, or
auction moments enter when a particular model requires them. Shares mean
*quantity* shares for Logit and *revenue* shares for CES and AIDS. A known
outside share or a normalization is essential when utilities or valuations
are identified only up to a common level.

For differentiated-product Bertrand pricing, the old reference manual's
core argument is still useful. Firm $f$ chooses prices for its products to
maximize $\sum_{j\in J_f}(p_j-c_j)q_j(p)$. Its first-order condition for
product $i$ is

$$
q_i(p)+\sum_{j\in J_f}(p_j-c_j)\frac{\partial q_j(p)}{\partial p_i}=0.
$$

Ownership weights generalize the sum to partial ownership. Observed margins
and shares restrict the demand derivatives in this condition. A calibrated
demand system supplies those derivatives; the same condition recovers
marginal costs. The counterfactual holds identified demand primitives fixed,
changes ownership or another environment field, and solves the appropriate
model's new equilibrium. Curvature away from the observed point therefore
matters even when two systems match the same baseline.

The quantity diversion from $i$ to $j$ is
$D^q_{ij}=-(\partial q_j/\partial p_i)/(\partial q_i/\partial p_i)$.
It helps identify substitution derivatives when shares and margins alone
do not. Its diagonal is $-1$ under the package's diversion-matrix convention;
the off-diagonal modeled products need not absorb all lost sales when there
is an outside option. See `?diversion` for the selected model's quantity or
revenue convention.

## Demand systems

The following compresses the economically useful demand descriptions from
the previous reference manual. Exact argument rules, solver controls, and
special cases live in the model's help page.

| Family | Demand shape and recovered information | Identification and limits |
|:--|:--|:--|
| Linear | $q=a+Bp$; product intercepts and a slope matrix give constant level derivatives. | Observed prices and quantities, margins, and diversion or symmetry restrictions identify otherwise underdetermined slopes. Counterfactual curvature is zero in levels. |
| Log-linear (`LogLin`) | $\log q_i=a_i+\sum_j B_{ij}\log p_j$; $B$ is an elasticity matrix. | Prices, quantities, margins, and diversion/restrictions identify slopes; own elasticities are constant rather than level slopes. |
| LA-AIDS (`AIDS`) | Budget shares vary with log prices and an expenditure/price index; cross-price restrictions encode adding up, homogeneity, and symmetry. | Revenue shares and observed margins plus diversion or curvature restrictions identify the local system; price levels may be optional for calibration but affect level outputs. |
| PCAIDS and nested PCAIDS | Proportional-calibration AIDS restricts substitution using share proportionality; nested PCAIDS relaxes substitution within nests. | `knownElast` and `mktElast` are explicit identifying inputs. Nested calibration also needs nests and sufficient margins; these are separate registry models, not arbitrary AIDS flags. |
| Logit | $s_i=\exp(\delta_i+\alpha p_i)/(1+\sum_j\exp(\delta_j+\alpha p_j))$ with $\alpha<0$ in output markets. | Prices, quantity shares, ownership, and enough margins identify the common price coefficient and mean values. With no outside option, one mean value is normalized. Logit imposes share-proportional substitution. |
| Nested Logit | Within-nest correlation changes diversion toward products in the same nest. | Nest membership and additional margin information identify nesting parameter(s); a singleton nest cannot identify its own parameter. The flat model is a restriction, not a new observed market. |
| CES | Revenue shares arise from quality weights and price powers; `gamma` controls substitution and `alpha`/inside share describes expenditure allocation. | Prices, revenue shares, margins and an outside share or normalization identify the relevant primitives under the selected construction. Ordinary output-market translations require `gamma > 1`. Price responses differ from Logit even at a matching baseline. |
| Nested CES | CES substitution differs within and across nests. | Nest assignments and nest curvatures require extra restrictions or margins. In the ordinary output-market domain, within-nest `sigma` exceeds between-nest `gamma > 1`; these are not nested-Logit parameters. |
| BLP/random coefficients | Heterogeneous price tastes are integrated over consumer draws, producing flexible substitution. | Observed prices, shares, known outside share `s0`, and conduct-specific margin information identify price-taste parameters by contraction and margin matching. Current calibration does not estimate `s0` or bargaining power. Supplied-parameter `specify()` uses the documented integration contract. |
| Capacity Logit (`LogitCap`) | Logit choice is combined with product quantity ceilings. | Market size converts shares to quantities; capacities and margins at exactly binding products matter because unconstrained FOCs do not identify their marginal costs. |

The Linear, Log-linear, and AIDS discussion follows the original treatment
of price derivatives, adding-up and diversion restrictions [@Haefen2002;
@LaFrance2004; @Epstein2004]. Logit, CES, and capacity cases retain the
original reference's distinction between substitution and supply-side
constraints [@Werden1994; @Sheu2011; @Froeb2003].

Unknown-outside-share calibrations are explicit `variant = "alm"` registry
entries where available. Additional margin information identifies the missing
outside share jointly with demand curvature. A variant with the same demand
name is still a distinct complete implementation and may lack `specify()`.

## Conduct and cost recovery

The demand family does not by itself specify firm behavior. The following
models solve distinct economic problems and recover costs with their own
conditions.

| Conduct | Strategic condition and merger interpretation | Calibration requirements |
|:--|:--|:--|
| Bertrand | Firms choose prices; ownership-weighted demand derivatives appear in the price FOCs above. A merger changes whose margins are internalized. | Observed margins help identify demand derivatives; costs follow from fitted margins and prices. |
| Cournot | Firms or plants choose quantities; each plant internalizes its effect on market inverse demand. Merger ownership changes quantity incentives. | Linear/LogLin general Cournot uses plant quantities and cost/production structure; Logit/CES Cournot have their own complete implementations. The Bertrand cost equation is not reused as a universal formula. |
| Stackelberg | Leaders choose quantities anticipating followers' response; merger, capacity and leader/product state can alter the sequence. | General Linear/LogLin quantity-game inputs include plant and leadership information. It is registered only for those implemented combinations. |
| Monopolistic competition (`moncom`) | Each product prices against its *perceived own* demand, without internalizing cross-product effects. CES holds its aggregate index fixed for the perceived derivative. | Prices, shares and at least one margin identify Logit/CES/BLP variants under their separate FOCs. Ownership is metadata, not a cross-product term in the MonCom FOC. |
| Differentiated second-score auction | Sellers offer the portfolio product with highest surplus and bid cost; payment reflects the next-best score. A merger changes the portfolio alternative and expected payment. | Shares and margin information recover offer-value and cost primitives. One valuation may require normalization; price data matter when decomposing value and cost for products with cost changes [@Brannman2000; @miller2014]. |
| Nash bargaining | Buyer and seller jointly determine a price against disagreement payoffs, accounting for recapture within the seller's portfolio. A merger changes both diversion and outside bargaining options. | Bargaining power is supplied, not inferred from an ordinary Bertrand margin; Logit/CES and supported BLP entries have their own FOCs. |
| Second-score bargaining | Bargaining uses a second-score auction disagreement environment. | Separate complete Logit/CES implementations combine their own demand, bargaining and auction states. |
| Capacity-constrained second-price auction | A buyer sets a reserve; sellers' costs are draws from a chosen distribution and capacities affect the distribution of the minimum cost. | Capacities, prices/margins and possibly reserve or self-supply moments identify distribution and reserve parameters; more flexible distributions need more independent moments [@WaehrP2003]. |

The standard second-score outcome and capacity-constrained auction are not
the same cost model. Likewise, a first-score differentiated procurement
setting may have a Bertrand interpretation under fixed non-price features,
but that observation does not create a new registry combination. Static
merger FOCs also differ from repeated-game coordinated-effects tools such
as `calcProducerSurplusGrimTrigger`; those remain separate legacy analyses,
not a `calibrate()` conduct entry.

## Implemented combinations

The table is generated at vignette-build time from the same registry as
`supportedModels()`. Each row is a **complete implementation**, not a
promise that any listed demand can be paired with any listed conduct. `TRUE`
means the lifecycle operation is implemented for that row. `simulate()`
still validates each counterfactual field against its model-specific
capabilities.

```{r support-table, echo=FALSE}
models <- antitrust::supportedModels()
knitr::kable(
  models[c("demand", "conduct", "variant", "class",
           "calibrate", "specify", "simulate")],
  row.names = FALSE
)
```

The registry also governs model identity and transition eligibility.
`respecify()` uses a *separate*, narrower transition registry; an implemented
row does not imply a conversion from every other row. See the [workflow
vignette](Workflow.html) for controlled transitions and the [architecture
vignette](Architecture.html) for dispatch and extensions.
