---
title: "Workflow and counterfactual analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow and counterfactual analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The working unit in `antitrust` is a **fitted structural market**. Fit observed
market data once; change ownership, costs, products, or other supported parts
of the environment; then solve the new equilibrium. The [models
vignette](Models.html) explains the economics. This vignette follows one
three-product Logit–Bertrand market throughout.

## Fit and inspect the observed market

The three products begin under separate owners. Observed prices, quantity
shares (with an outside option), and margins identify Logit demand under
Bertrand pricing. `calibrate()` delegates the economic inversion and cost
recovery to the registered Logit model.

```{r baseline}
library(antitrust)

prices <- c(2, 2.2, 2.5)
shares <- c(.35, .25, .20)
margins <- c(.40, .35, .30)
owners <- c("A", "B", "C")

fit <- calibrate(
  "logit", "bertrand", prices = prices, shares = shares,
  margins = margins, ownerPre = owners, insideSize = 100
)
summary(fit)[c("spec", "model_class")]
fit@parameters$alpha
```

An `AntitrustFit` stores the model specification, the fitted legacy S4 model,
recovered parameters, observed inputs, and diagnostics. `summary(fit)` reads
that stored state; it does not solve a new equilibrium. The baseline result is
available in `fit@model`, including its `pricePre` and `mcPre` slots.

## Ask a merger question

`counterfactual()` describes changes relative to the fit. `simulate()` solves
one equilibrium and returns the established `Logit` result class, so its
ordinary methods still apply.

```{r merger}
merger <- counterfactual(ownership = c("A", "A", "C"))
merged <- simulate(fit, merger)
data.frame(before = merged@pricePre, after = merged@pricePost)
```

The fit is reusable. These are distinct *one-step* questions about the same
observed market, not sequential changes to `merged`:

```{r alternatives}
efficient_merger <- simulate(fit, counterfactual(
  ownership = c("A", "A", "C"), costs = c(0, -.10, 0)
))
exit_result <- simulate(fit, counterfactual(exit = "Prod3"))
entrant_result <- simulate(fit, counterfactual(
  entry = entrant("New", meanval = 0, cost = 1, priceStart = 2)
))

data.frame(merger = merged@pricePost,
           merger_with_efficiency = efficient_merger@pricePost)
exit_result@pricePost
entrant_result@pricePost
```

Here `costs = -.10` is a 10% reduction for product 2 under the
constant-cost Logit–Bertrand convention. `exit` removes a product from the
solved market. An entrant supplies its own label, mean value, marginal cost,
and solver starting price; it inherits the fitted market's common Logit price
coefficient. `entry` is supported only for registered model classes with an
entry implementation. A named `quality` shock is another supported Logit
scenario: for example, `quality = c(Prod1 = .05)` increases product 1's
choice weight by 5%. The original `fit` remains unchanged.

Changes in a single `counterfactual()` are simultaneous. Separate one-step
objects can be combined when their fields do not conflict:

```{r combine}
combined <- combine_counterfactuals(
  merger, counterfactual(costs = c(0, -.10, 0))
)
combined_result <- simulate(fit, combined)
combined_result@pricePost
```

`combined_result` answers the same simultaneous question as
`efficient_merger`. Conflicting values for the same field are rejected.

## Sequence changes when order matters

`add_step()` adds a later equilibrium solve. Each step starts from the
previous step's solved state; costs and product state persist according to
the fitted model's rules. A multi-step simulation returns a
`CounterfactualPath`, with a result for each step.

```{r path}
staged <- add_step(merger, costs = c(0, -.10, 0))
path <- simulate(fit, staged)
data.frame(after_merger = result_at(path, 1)@pricePost,
           after_efficiency = final_result(path)@pricePost)
```

`result_at(path, 1)` is the intermediate `Logit` result;
`final_result(path)` is the last one. A path can be resumed with
`simulate(path, counterfactual(...))`. See `?add_step` and `?simulate` for
the persistence contract.

## Choose the right operation

| Operation | Economic question | Effect on fitted parameters |
|:--|:--|:--|
| `calibrate()` | What structure is identified by observed prices, shares, margins and a chosen economic model? | Recovers demand primitives and costs using that model's calibration equations. |
| `specify()` | What if structural parameters are already known or supplied? | Constructs the supported model from those parameters; no demand recalibration. |
| `simulate()` | What equilibrium follows from a changed economic environment? | Keeps the fit's structural demand; applies the supported scenario and solves. |
| `update()` | What if an observed calibration input changes? | Re-runs calibration for the **same** specification. |
| `respecify()` | What can be carried into another registered model? | Uses an explicit transition, reconstructing target state without calibrating from the source margins. |

For this fit, revising an observed margin is a new calibration. Supplying
known Logit parameters instead is a different construction route:

```{r calibration-routes}
revised <- update(fit, margins = c(.38, .35, .30))
known <- specify(
  "logit", "bertrand", prices = prices,
  ownerPre = owners, parameters = fit@parameters, insideSize = 100
)
c(revised_alpha = revised@parameters$alpha,
  supplied_alpha = known@parameters$alpha)
```

`update()` currently accepts replacement observed inputs and calibration
options for the fit's *same* demand, conduct, and variant. A changed model
specification needs its own `calibrate()` call if fresh identification is the
question. A fit created by `specify()` has no observed-data calibration call
to update.

## Translate a fitted market, with limits

`respecify()` is an explicitly registered transition, never arbitrary S4
coercion. For the fitted market, a same-demand move from Bertrand to Cournot
retains portable Logit demand primitives and rebuilds the supply and marginal
cost state under Cournot. A Logit-to-CES translation requires a target CES
curvature (`gamma`); it holds baseline prices, shares/quantities, ownership,
and active products fixed, solves target mean values, and recovers target
costs. The source Logit price coefficient and source cost state cease to be
valid CES primitives.

```{r transitions, warning=FALSE}
cournot_fit <- respecify(fit, conduct = "cournot")
ces_fit <- respecify(fit, demand = "ces", gamma = 2)
linear_fit <- respecify(fit, demand = "linear")
vapply(list(cournot_fit, ces_fit, linear_fit),
       function(x) x@spec$id, character(1))
```

Flat/nested Logit and CES transitions are also registered when the required
nest assignments and curvature are supplied; nested-to-flat transitions
discard the nesting primitive. First-order Linear and LogLin transitions
match the fitted baseline and its local derivatives. They do not preserve
global demand shape. Inspect a proposed transition with
`model_transition(model_spec("logit", "bertrand"),
model_spec("ces", "bertrand"))` before applying it.

Matching a baseline does **not** make Logit and CES globally equivalent:
their substitution and curvature away from that point differ, so merger
predictions can differ. Translation diagnostics are recorded in
`ces_fit@diagnostics$translation`. `respecify()` deliberately clears the
target's recalibration call: `update(ces_fit)` cannot silently identify CES
from the original Logit margins.

## Scope of counterfactual fields

The scenario vocabulary is shared, but each registered implementation
accepts only fields it can solve. `capacity` is public for `LogitCap` and
Stackelberg models; `bargaining` applies to bargaining models; `leader` and
`products` apply to Stackelberg. Their calibration requirements and cost
conventions differ from this Logit–Bertrand example. `tariff` and `quota`
are reserved for the sibling `trade` package, and are rejected by antitrust
models. `?counterfactual`, `?supportedModels`, and the [models
vignette](Models.html) give the model-specific boundaries.
