---
title: "Workflow and counterfactual analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow and counterfactual analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Suppose firms A and B propose a merger. You want to know how their prices
would change, whether a cost saving would offset that change, and how the
answer depends on which products remain in the market. These questions share
the same observed market. In `antitrust`, you fit that market once and then
change the conditions under which firms compete.

We will use one three-product Logit–Bertrand market throughout. The [models
vignette](Models.html) explains the economic assumptions behind the fit.

## Explore an observed synthetic market

Structural demand parameters can be difficult to choose directly. Observed
synthetic mode is an intuition/calibration experiment: choose shares,
ownership, marginal costs, and one reference proportional margin, then let
the selected model imply its demand parameter and equilibrium prices. It is
not an empirical data-generating process.

```{r synthetic-observed}
intuition <- antitrust::synthetic_market(
  demand = "logit", supply = "bertrand",
  n_firms = 2, n_products = 2,
  shares = c(.2, .1, .15, .25, .3),
  costs = c(40, 50, 60, 70, 80),
  reference_margin = .2,
  seed = 7
)
intuition@diagnostics$synthetic[c(
  "supplied_costs", "solved_prices", "recovered_parameters",
  "share_residual", "foc_residual")]
```

The reference product has cost 80 and margin 0.2, so its implied price is
`80/(1-0.2) = 100`. The other prices follow the ownership-adjusted Bertrand
FOCs. A different demand or conduct specification must satisfy its own
equations; see the [models vignette](Models.html) and
`supportedModels()$observed_synthetic` for supported routes.

For ALM, `passive_outside_share` describes a separate outside option. The
product shares above remain conditional on choosing an active product;
omitting this argument draws it uniformly from 0.1 to 0.5. For example,
`synthetic_market(demand = "LogitALM", supply = "cournot",
shares = c(.2, .1, .15, .25, .3), costs = c(40, 50, 60, 70, 80),
n_firms = 2, n_products = 2, reference_margin = .2,
passive_outside_share = .3)` solves the Cournot equation with that passive
share and the same known costs.

## Fit and inspect the observed market

Each firm initially sells one product. The products account for 80% of
quantity, leaving 20% for the outside option. We observe their prices and
gross margins as well as their shares. Under the chosen Logit demand and
Bertrand pricing assumptions, `calibrate()` uses those observations to recover
demand parameters and marginal costs.

```{r baseline}
library(antitrust)

prices <- c(2, 2.2, 2.5)
shares <- c(.35, .25, .20)
margins <- c(.40, .35, .30)
owners <- c("A", "B", "C")

fit <- calibrate(
  "logit", "bertrand", prices = prices, shares = shares,
  margins = margins, ownerPre = owners, insideSize = 100
)
summary(fit)[c("spec", "model_class")]
fit@parameters$alpha
```

The returned `AntitrustFit` holds the observations, recovered parameters,
model specification, and fitted model. `summary(fit)` describes this fitted
state; it does not predict a new outcome. The underlying model is in
`fit@model`, where `pricePre` and `mcPre` hold baseline prices and costs.

## Ask a merger question

If A acquires B, the merged firm takes account of sales that move between
their products when it changes either price. Represent that change with a
`counterfactual()`, then use `simulate()` to solve for postmerger prices.

```{r merger}
merger <- counterfactual(ownership = c("A", "A", "C"))
merged <- simulate(fit, merger)
data.frame(before = merged@pricePre, after = merged@pricePost)
```

The result is a `Logit` object, so the familiar model methods remain
available. We can use the same fit to ask several other questions. For
example, what if the merger reduces product 2's marginal cost? What if
product 3 exits, or a new product enters? Each simulation below starts from
the **original fitted market**, regardless of the simulations above it.

```{r alternatives}
efficient_merger <- simulate(fit, counterfactual(
  ownership = c("A", "A", "C"), costs = c(0, -.10, 0)
))
exit_result <- simulate(fit, counterfactual(exit = "Prod3"))
entrant_result <- simulate(fit, counterfactual(
  entry = entrant("New", meanval = 0, cost = 1, priceStart = 2)
))

data.frame(merger = merged@pricePost,
           merger_with_efficiency = efficient_merger@pricePost)
exit_result@pricePost
entrant_result@pricePost
```

In this model, `costs = -.10` means a 10% reduction in product 2's marginal
cost. `exit` removes product 3 from the solved market. The entrant needs its
own label, mean value, marginal cost, and starting price for the solver; it
shares the fitted Logit price coefficient. This entry calculation is available
only for models with an entry implementation. A named quality change is
possible too: `quality = c(Prod1 = .05)` raises product 1's choice weight by
5%. None of these questions changes `fit`.

The merger and cost saving above occur together. You can also define them
separately and combine them before solving:

```{r combine}
combined <- combine_counterfactuals(
  merger, counterfactual(costs = c(0, -.10, 0))
)
combined_result <- simulate(fit, combined)
combined_result@pricePost
```

`combined_result` answers the same question as `efficient_merger`.
`combine_counterfactuals()` rejects conflicting values for the same field.

## See what happens at each stage

Suppose the firms merge before they realize a cost saving. The final prices
are only part of the analysis; you may also want prices in the period between
the two changes. `add_step()` creates a second solve, starting from the
outcome of the first. Costs and product state carry forward under the fitted
model's rules. The result is a `CounterfactualPath` containing both
equilibria.

```{r path}
staged <- add_step(merger, costs = c(0, -.10, 0))
path <- simulate(fit, staged)
data.frame(after_merger = result_at(path, 1)@pricePost,
           after_efficiency = final_result(path)@pricePost)
```

`result_at(path, 1)` retrieves the outcome after the merger, and
`final_result(path)` retrieves the outcome after the cost saving. Both are
`Logit` results. You can continue a path with
`simulate(path, counterfactual(...))`; `?add_step` and `?simulate` describe
which changes carry forward.

## Which question are you asking?

The five lifecycle functions answer different questions. In particular,
new evidence about the observed market calls for a new calibration, while a
proposed change to that market calls for a simulation.

| Operation | Use it when... | What happens |
|:--|:--|:--|
| `calibrate()` | You have observed market inputs and an economic model. | Recovers demand parameters and costs using that model's calibration equations. |
| `specify()` | You already know the structural parameters. | Builds a fit from supplied parameters, without recalibrating demand. |
| `simulate()` | You want the equilibrium after a proposed change. | Applies the change to a fit and solves under its structural model. |
| `update()` | An observed input or calibration option has changed. | Calibrates again under the **same** model specification. |
| `respecify()` | You want to examine another supported model specification. | Transfers only the information allowed by a registered transition and reconstructs the target model. |

For example, a revised estimate of product 1's observed margin changes what
the data identify. It calls for `update()`. If Logit parameters are already
known, `specify()` provides another way to construct a fit:

```{r calibration-routes}
revised <- update(fit, margins = c(.38, .35, .30))
known <- specify(
  "logit", "bertrand", prices = prices,
  ownerPre = owners, parameters = fit@parameters, insideSize = 100
)
c(revised_alpha = revised@parameters$alpha,
  supplied_alpha = known@parameters$alpha)
```

`update()` accepts revised inputs or calibration options for the fit's
existing demand, conduct, and variant. To identify a different model from
observations, call `calibrate()` with that model. A fit made by `specify()`
has no observed-data calibration call to repeat.

## Translate a fitted market, with limits

Different assumptions can fit the same observed market yet predict different
responses to a merger. `respecify()` lets us explore certain comparisons
without treating one economic model as interchangeable with another. It
works only where the package has a registered transition.

Moving this Logit market from Bertrand to Cournot carries over its portable
demand parameters, then rebuilds the supply and marginal-cost state under
Cournot behavior. Moving from Logit to CES needs a target CES curvature
(`gamma`). That transition preserves baseline prices, shares or quantities,
ownership, and active products; it solves for CES mean values and recovers
costs under the target model. The Logit price coefficient and recovered costs
are no longer valid CES parameters.

```{r transitions, warning=FALSE}
cournot_fit <- respecify(fit, conduct = "cournot")
ces_fit <- respecify(fit, demand = "ces", gamma = 2)
linear_fit <- respecify(fit, demand = "linear")
vapply(list(cournot_fit, ces_fit, linear_fit),
       function(x) x@spec$id, character(1))
```

Transitions between flat and nested Logit or CES models require the relevant
nest assignments and curvature; going back to a flat model discards the
nesting parameter. First-order translations to Linear or LogLin match the
baseline and its local derivatives, rather than demand far from that point.
Inspect a proposed transition with
`model_transition(model_spec("logit", "bertrand"),
model_spec("ces", "bertrand"))` before applying it.

Thus a Logit-to-CES transition is a **local, baseline-preserving comparison**,
not a claim that the two demand systems are globally equivalent. Their
substitution and curvature away from the baseline differ, and so may their
merger predictions. Translation diagnostics are in
`ces_fit@diagnostics$translation`. The translated fit has no CES calibration
call: `update(ces_fit)` cannot quietly recalibrate CES from the original
Logit margins.

## Scope of counterfactual fields

Not every model can answer every scenario. Capacity changes are available for
`LogitCap` and Stackelberg models; bargaining changes belong to bargaining
models; leader and product-state changes belong to Stackelberg models. These
models need different observations and cost assumptions from our
Logit–Bertrand example. Tariffs and quotas belong to the sibling `trade`
package. Consult `?counterfactual`, `?supportedModels`, and the [models
vignette](Models.html) before applying a field to another model.
